---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
```{r message=FALSE}
library(dplyr)

if (!file.exists("./activity.csv")) {
        unzip("./activity.zip")
}

dataset <- read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"))
print(head(dataset))
```



## What is mean total number of steps taken per day?

### Computation of the number of steps taken each day

```{r message=FALSE}

total_steps_per_day <- dataset %>%
  as_tibble %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = TRUE))

print(head(total_steps_per_day))
```

### Histogram of the total number of steps taken each day

```{r}
library(lattice)

histogram(~ total_steps, data = total_steps_per_day, 
          breaks = 20,
          xlab = "Total Steps",
          main = "Histogram of the total numbers of steps taken each day")
```

### Mean and median of the total number of steps taken each day

```{r}
mean <- mean(total_steps_per_day$total_steps)
median <- median(total_steps_per_day$total_steps)
```
The mean of the total number of steps taken each day is **`r mean`**.

The median of the total number of steps taken each day is **`r median`**.

## What is the average daily activity pattern?

### Computation of the average steps for each interval

```{r message = FALSE}
average_steps_by_interval <- dataset %>%
  as_tibble %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps, na.rm = TRUE))
```

### Time series plot of the 5-minute interval
```{r message=FALSE}
library(lattice)
xyplot(mean_steps ~ interval, type = "l", data = average_steps_by_interval,
       main = "Time series plot of the 5-minute interval",
       xlab = "Interval",
       ylab = "Average steps")
```

```{r}
max_steps <- max(average_steps_by_interval$mean_steps)
interval_for_max_steps <- average_steps_by_interval %>%
  filter(mean_steps == max(max_steps)) %>%
  .$interval
```

The interval **`r interval_for_max_steps`** has the maximum numbers of steps **`r max_steps`**

## Imputing missing values

### Total number of missing values in the dataset
```{r}
number_of_missing_values <- sum(!complete.cases(dataset))
```

There are **`r number_of_missing_values`** missing values in the dataset.

### Imputing strategy

Missing values are replaced by the mean of steps for the corresponding intervals

```{r}
imputed_dataset <- dataset %>% group_by(interval) %>% mutate(steps=ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))

print(head(imputed_dataset))
```

```{r include=FALSE}
imputed_dataset <- dataset
imputed_dataset$steps <- ave(imputed_dataset$steps, imputed_dataset$interval, FUN = function(x) {ifelse(is.na(x), mean(x, na.rm = TRUE), x)})

print(head(imputed_dataset))
```

### Histogram of the total number of steps taken each day

```{r message=FALSE}
imputed_total_steps_per_day <- imputed_dataset %>%
  as_tibble %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = TRUE))

histogram(~ total_steps, data = imputed_total_steps_per_day,
          breaks = 20,
          xlab = "Total Steps",
          main = "Histogram of the total numbers of steps taken each day")

imputed_mean <- mean(imputed_total_steps_per_day$total_steps)
imputed_median <- median(imputed_total_steps_per_day$total_steps)

```

The mean of of the total number of steps for the imputed dataset is **`r imputed_mean`**

The median of of the total number of steps for the imputed dataset is **`r imputed_median`**

## Are there differences in activity patterns between weekdays and weekends?

### Factor variable telling if a date is a weekend or weekday

```{r}
weekday_dataset <- imputed_dataset %>%
  mutate(weekday = ifelse(as.POSIXlt(date)$wday %in% c(0, 5), "weekend", "weekday"))

print(head(weekday_dataset))
```

### Time series plot for week days and week ends
```{r message = FALSE}
averaged_weekday_steps <- weekday_dataset %>%
  group_by(weekday, interval) %>%
  summarize(average_steps = mean(steps))

xyplot(average_steps ~ interval | weekday, type = "l", data = averaged_weekday_steps,
       xlab = "Interval",
       ylab = "Average steps",
       layout = c(1, 2))
```
